---
title: "Getting started with midasINLA"
output:
  rmarkdown::html_vignette:
    math_method: mathjax
vignette: >
  %\VignetteIndexEntry{Getting started with midasINLA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(midasINLA)
library(INLA)
library(ggplot2)
library(dplyr)
library(tidyr)
INLA::inla.setOption(num.threads = 1)
```


## Introduction

`midasINLA` provides tools for fitting mixed-frequency time-series models using the Integrated Nested Laplace Approximation (INLA) framework, with support for both constant and spatially varying regression coefficients.

The package allows high-frequency covariates to be incorporated into a
lower-frequency response model through MIDAS lag-weight functions. Different
constraint schemes can be used to model the lag weights, while coefficients
can be either constant or spatially varying.

## Model 

Consider a response variable $y_{it}$, indexed by spatial unit $i=1,\ldots,N$ and low-frequency time point $t=1,\ldots,T$. The predictor $x_{i\tau}$ is observed at a higher frequency. The MIDAS framework relates the low-frequency response to multiple high-frequency observations through a weighted distributed lag:

$$
\begin{aligned}
        &y_{it} \sim F, \;\;\; \mathbb{E}(y_{it}) = \mu_{it} \\
        &g(\mu_{it}) = \beta_0 + \beta_i\sum_{k=0}^K w_kx_{i,s(t)-k} + \epsilon_{it} \\
        &w_k=h(\boldsymbol{\gamma},k) \; \text{and} \; \sum_{k=0}^K w_k=1.
\end{aligned}
$$

Here, $g(\cdot)$ is the link function, $\beta_0$ is the intercept, and $\beta_i$ is the regression coefficient for the high-frequency predictor for the $i^{\text{th}}$ spatial unit. The function $h(\boldsymbol{\gamma},k)$ determines the MIDAS lag weights, where $\boldsymbol{\gamma}\in\mathbb{R}^d$ contains the parameters governing the shape of the weighting function. The weights are constrained to sum to one, which separates the overall magnitude of the predictor effect, represented by $\beta_i$, from the relative contributions of the individual lags.

The index $s(t)$ denotes the cumulative number of high-frequency observations up to low-frequency time point t,

$$
s(t)=\sum_{j=1}^{t}m_j,
$$

where $m_t$ is the number of high-frequency observations associated with the $t^{\text{th}}$ low-frequency observation. In the examples below, the same high-frequency sampling structure is assumed across spatial units.

The main flexibility of the model comes from the choice of the lag-weight function $h(\boldsymbol{\gamma},k)$. midasINLA provides functions for constructing different MIDAS weighting schemes and incorporating them into an INLA model, while allowing the regression coefficient to be either constant or spatially varying.

### Spatially varying coefficients

midasINLA allows the regression coefficient $\beta_i$ to vary across spatial units in two ways:

* Spatially structured coefficient:
    $$
    \beta_i = \beta^* + b_i, \qquad
    \boldsymbol{b} \sim \operatorname{iCAR}(\tau_b,\mathbf{W}),
    $$
    where $\beta^*$ is the overall effect and $b_i$ represents the spatially structured deviation for unit $i$. The deviations follow an intrinsic conditional autoregressive (iCAR) model with precision $\tau_b$ and spatial weights matrix $\mathbf{W}$, subject to the sum-to-zero constraint $\sum_i b_i=0$.
* Unstructured coefficient:
    $$
    \beta_i \overset{iid}{\sim}
    N(0,\sigma_\beta^2),
    $$
    where the coefficients are independent across spatial units.

A constant coefficient, $\beta_i\equiv\beta$, is also supported and corresponds to the special case in which the effect is the same across all spatial units.


### Constraint functions

The lag weights are obtained by normalising a constraint function $\psi(\boldsymbol{\gamma},k)$:

$$
w_k = h(\boldsymbol{\gamma},k)
= \frac{\psi(\boldsymbol{\gamma},k)}
{\sum_{j=0}^K \psi(\boldsymbol{\gamma},j)}.
$$

midasINLA implements several commonly used constraint functions.

1. Exponential Almon polynomial (order 2):

$$
\psi(\boldsymbol{\gamma},k) = \exp\left(\sum_{j=1}^{2}\gamma_j k^j\right),
$$

where $\boldsymbol{\gamma}=(\gamma_1,\gamma_2)$.

2. Beta polynomial:

$$
\psi(\boldsymbol{\gamma},k) = x_k^{\gamma_1-1}(1-x_k)^{\gamma_2-1},
$$

where

$$
x_k=\xi+(1-2\xi)\frac{k}{K},
$$

with $\xi>0$ a small fixed constant and $\boldsymbol{\gamma}=(\gamma_1,\gamma_2)$. A one-parameter version is obtained by fixing $\gamma_1=1$.

3. Hyperbolic scheme:

$$
\psi(\gamma,k) = \frac{\Gamma(k+\gamma)}{\Gamma(k+1)\Gamma(\gamma)},
$$
where $\gamma>0$.

4. Gaussian kernel:

$$
\psi(\boldsymbol{\gamma},k)=\exp\left\{-\frac{(k-\gamma_1)^2}{2\gamma_2}\right\}.
$$

where $\boldsymbol{\gamma}=(\gamma_1,\gamma_2)$ and $\gamma_2>0$.


## INLA implementation

The MIDAS lag structure is incorporated into the latent Gaussian model through INLA’s `rgeneric` interface. The MIDAS constraint functions define the lag weights as a function of a low-dimensional parameter vector, while the resulting weighted high-frequency covariates are represented as part of the latent model.

The functions in `midasINLA` construct the required `rgeneric` model components and interface them with `INLA::inla()`. This allows the MIDAS lag-weight parameters and regression coefficients to be estimated within the INLA framework, while retaining the spatial structure specified for the regression coefficients.

Users do not need to construct the `rgeneric` model directly; this is handled internally by the package functions demonstrated below.


## Simulated spatial Poisson example

We show an example with two high-frequency covariates with
different lag-weight constraints. The first covariate has a spatially
varying coefficient, whereas the second has a constant coefficient.

The example illustrates the main functions in midasINLA:

* `prepare_Minla_spatial`() prepares high-frequency covariates for inclusion
    in the model;
* `fit_Minla_spatial`() fits the resulting model using INLA;
* `compute_beta_spatial`() obtains posterior summaries of the regression
    coefficients;
* `compute_weights`() obtains posterior summaries of the MIDAS lag weights;
* `predict_midas`() generates posterior predictions.

We consider an outcome $y_{it}$ observed at 16 spatial locations and
192 time points. Two high-frequency covariates, $x_{1it}$ and $x_{2it}$,
are available for each location. There are 30 high-frequency observations
corresponding to each response time point.

The first covariate uses a hyperbolic lag-weight constraint with
$\gamma = 0.9$ and 29 lags. Its regression coefficient varies spatially
according to an intrinsic conditional autoregressive (iCAR) model.

The second covariate uses a Gaussian lag-weight constraint with
$\gamma_1 = 10$ and $\sqrt{\gamma_2} = 12$ and 45 lags. Its regression
coefficient is constant across locations.

The simulated data are included with the package and can be loaded using:

```{r load-data}
data("data_spatialpoisson_example")
```

The dataset is provided as a list containing the response, two high-frequency covariates, the spatial polygons, and the true parameter values used to generate the data. The available components can be inspected with:

```{r inspect-data}
names(data_spatialpoisson_example)
```

The response data are stored in data_y and contain the outcome together with the spatial and temporal indices:

```{r response-data}
head(data_spatialpoisson_example[["data_y"]])
```

The high-frequency covariates are stored in `data_x1` and `data_x2`. Each contains the covariate values together with their corresponding spatial indices:

```{r covariate-data}
head(data_spatialpoisson_example$data_x1)
head(data_spatialpoisson_example$data_x2)
```

The neighbourhood structure used to generate the spatially varying
coefficient is stored as an `inla.graph` object. It can be loaded from
the package using:

```{r graph}
g <- INLA::inla.read.graph(
  filename = system.file("map.adj", package = "midasINLA")
)
```

The data were generated using a Poisson model of the form

$$
y_{it} \sim \operatorname{Poisson}(\mu_{it}),
$$

with

$$
\log(\mu_{it}) =
\beta_0 +
(\beta_1^* + b_i)
\sum_{k=0}^{29} w_{1k}x_{1i,s(t)-k}
+
\beta_2
\sum_{k=0}^{45} w_{2k}x_{2i,s(t)-k},
$$

where $\boldsymbol{b}$ follows an iCAR model. The true values of the
parameters are $\beta_0 = 1$, $\beta_1^* = 1.1$, and $\beta_2 = -2$.


## Preparing the MIDAS covariates

Before fitting the model, each high-frequency covariate is prepared using
`prepare_Minla_spatial`().

For the first covariate, we use the hyperbolic constraint and allow the
coefficient to vary spatially according to an iCAR model:


```{r prepare-x1}
Midas_x1 <- prepare_Minla_spatial(
  x = data_spatialpoisson_example$data_x1$x1,
  loc_x = data_spatialpoisson_example$data_x1$loc,
  constraint = "hyperbolic",
  K = 0:29,
  m = 30,
  svc = TRUE,
  svc_prior = "icar",
  g = g
)
```

```{r prepare-x2}
Midas_x2 <- prepare_Minla_spatial(
  x = data_spatialpoisson_example$data_x2$x2,
  loc_x = data_spatialpoisson_example$data_x2$loc,
  constraint = "gaussian",
  K = 0:45,
  m = 30,
  svc = FALSE
)
```

The resulting objects contain the information required by
`fit_Minla_spatial`() to construct the MIDAS components of the model.


## Fitting the model

The response data are stored in data_y. To illustrate prediction, we
reserve the final 10 response time points at each location as a test set.

```{r prepare-response}
response_data <- data_spatialpoisson_example[["data_y"]]

response_data$y_all <- response_data$y

response_data[which(response_data[["Time"]] %in% 183:192),"y"] <- NA
```

The model can then be fitted using `fit_Minla_spatial`():
  
```{r fit-model, eval = TRUE}
fit_res <- fit_Minla_spatial(
  formula = y ~ 1,
  data = response_data,
  loc_var = "loc",
  time_var = "Time",
  family = "poisson",
  hf_input = list(Midas_x1, Midas_x2),
  inla_options = list(verbose = FALSE,
                      control.predictor = list(
                        compute = TRUE,link = 1)))
```


A summary of the fitted model can be obtained using the standard
`summary`() method:
  
```{r model-summary}
summary(fit_res[["res"]])
```

The fitted model contains two MIDAS components, corresponding to the two
high-frequency covariates. The model also estimates the spatially varying
coefficient associated with the first covariate. The parameters governing
the MIDAS lag-weight functions are reported with names beginning with
`hf_idx_`.


## Posterior summaries of regression coefficients

Posterior summaries of the MIDAS regression coefficients can be obtained
using `compute_beta_spatial`().

```{r beta-results}
beta_results <- compute_beta_spatial(
  model = fit_res,
  n_loc = 16
)
```

The returned object contains results for each high-frequency covariate,
including marginal distributions and posterior summaries.

For the first covariate, the posterior summaries of the spatially varying
component $b_i$ can be accessed using:
  
```{r beta-summary}
beta_results$hf_index_1$summary.icar.beta
```

The spatially varying coefficient at location $i$ is defined as

$$
  \beta_{1,i} = \beta_1^* + b_i.
$$
Posterior summaries of the resulting total coefficient can be accessed using:

```{r total-beta-summary}
beta_results$hf_index_1$summary.total.beta
```

For the second covariate, which has a constant regression coefficient,
posterior summaries of $\beta_2$ are available using:

```{r constant-beta-summary}
beta_results$hf_index_2$summary.beta
```

## Estimating the MIDAS lag weights

The posterior distributions of the lag weights can be obtained using
`compute_weights`():

```{r weights}
res_weights <- compute_weights(fit_res)
```

The result is a list containing one data frame for each high-frequency
covariate. Each data frame contains the lag, posterior mean, and lower and
upper posterior quantiles.

For example, the estimated weights for the first covariate are:

```{r weights-x1}
head(res_weights$hf_1)
```

The weights for the second covariate can be inspected similarly:

```{r weights-x2}
head(res_weights$hf_2)
```

The posterior summaries can be used to visualise the estimated
lag-weight functions. Because the data are simulated, the true lag
weights are also available for comparison.

```{r plot-weights, fig.width = 7, fig.height = 4.5}
ggplot(res_weights$hf_1, aes(x = lag, y = mean)) +
  geom_errorbar(
    aes(
      ymin = q2.5,
      ymax = q97.5
    ),
    width = 0.2,
    colour = "grey30"
  ) +
  geom_point(
    aes(
      colour = "Posterior mean"
    ),
    size = 2
  ) +
  geom_point(
    aes(
      y = data_spatialpoisson_example$weights1,
      colour = "True value"
    ),
    size = 2
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Posterior mean" = "red",
      "True value" = "blue"
    )
  ) +
  labs(
    x = "Lag",
    y = "Lag weight"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  )
```

The corresponding lag-weight function for the second covariate can be
visualised in the same way:

```{r plot-weights-x2, fig.width = 7, fig.height = 4.5}
ggplot(res_weights$hf_2, aes(x = lag, y = mean)) +
  geom_errorbar(
    aes(
      ymin = q2.5,
      ymax = q97.5
    ),
    width = 0.2,
    colour = "grey30"
  ) +
  geom_point(
    aes(
      colour = "Posterior mean"
    ),
    size = 2
  ) +
  geom_point(
    aes(
      y = data_spatialpoisson_example$weights2,
      colour = "True value"
    ),
    size = 2
  ) +
  scale_colour_manual(
    name = NULL,
    values = c(
      "Posterior mean" = "red",
      "True value" = "blue"
    )
  ) +
  labs(
    x = "Lag",
    y = "Lag weight"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  )
```


## Prediction for held-out observations

The final 10 time points were withheld from the model fit and are used here
to illustrate posterior prediction.

Posterior predictions can be generated using `predict_midas()`:

```{r prediction}
pred_res <- predict_midas(
  model = fit_res,
  family = "poisson",
  Ntrials = NULL,
  nsamples = 1000
)
```

The returned object contains posterior summaries of the predicted outcome,
including posterior means and 95% credible intervals, as well as posterior
samples of the latent predictor. For example:
  
```{r prediction-structure}
str(pred_res, max.level = 2)
```
The posterior summaries of the predicted outcome can be accessed using:

```{r prediction-summary}
head(pred_res$computed_y$mean)
head(pred_res$computed_y$q2.5)
head(pred_res$computed_y$q97.5)
```

The predicted and observed outcomes can also be compared graphically. The
following example shows the results for the first four spatial locations.
The dashed vertical line indicates the boundary between the training and
held-out prediction periods.


```{r prediction-plot, fig.width = 7, fig.height = 6}
plot_data <- data.frame(
  observed = fit_res$data_final$y_all,
  predicted = pred_res$computed_y$mean,
  lower = pred_res$computed_y$q2.5,
  upper = pred_res$computed_y$q97.5,
  loc = fit_res$data_final$loc,
  Time = fit_res$data_final$Time
)

plot_long <- plot_data |>
  dplyr::filter(loc %in% 1:4) |>
  tidyr::pivot_longer(
    cols = c(observed, predicted),
    names_to = "series",
    values_to = "value"
  )

# Determine the training/held-out boundary for each location
non_na <- !is.na(fit_res$data_final$y)

segment <- cumsum(
  non_na != dplyr::lag(non_na, default = TRUE)
)
segment[!non_na] <- NA

rel_idx <- ave(
  seq_along(non_na),
  segment,
  FUN = seq_along
)

first_na <- which(
  diff(c(FALSE, is.na(fit_res$data_final$y))) == 1
)

vlines <- data.frame(
  cut = rel_idx[first_na - 1] + 1,
  loc = seq_len(16)
) |>
  dplyr::filter(loc %in% 1:4)

ggplot(plot_long, aes(x = Time, y = value, colour = series)) +
  geom_ribbon(
    data = plot_data |>
      dplyr::filter(loc %in% 1:4),
    aes(
      x = Time,
      ymin = lower,
      ymax = upper
    ),
    inherit.aes = FALSE,
    fill = "red",
    alpha = 0.2
  ) +
  geom_line() +
  geom_vline(
    data = vlines,
    aes(xintercept = cut),
    colour = "black",
    linetype = "dashed"
  ) +
  facet_wrap(
    ~loc,
    ncol = 2,
    labeller = labeller(
      loc = function(x) paste("Loc =", x)
    )
  ) +
  scale_colour_manual(
    values = c(
      "observed" = "blue",
      "predicted" = "red"
    )
  ) +
  labs(
    x = "Time",
    y = "Outcome",
    colour = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```
