---
title: "Getting Started with midasINLA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with midasINLA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(midasINLA)
library(INLA)

INLA::inla.setOption(num.threads = 1)
```


## Introduction

`midasINLA` provides tools for fitting mixed-frequency time-series models using the Integrated Nested Laplace Approximation (INLA) framework, with support for both constant and spatially varying regression coefficients.

The package allows high-frequency covariates to be incorporated into a
lower-frequency response model through MIDAS lag-weight functions. Different
constraint schemes can be used to model the lag weights, while coefficients
can be either constant or spatially varying.

This vignette demonstrates the main workflow using a simulated spatial
Poisson example. The example contains two high-frequency covariates with
different lag-weight constraints. The first covariate has a spatially
varying coefficient, whereas the second has a constant coefficient.

The example illustrates the main functions in midasINLA:

* `prepare_Minla_spatial`() prepares high-frequency covariates for inclusion
    in the model;
* `fit_Minla_spatial`() fits the resulting model using INLA;
* `compute_beta_spatial`() obtains posterior summaries of the regression
    coefficients;
* `compute_weights`() obtains posterior summaries of the MIDAS lag weights;
* `predict_midas`() generates posterior predictions.


## Simulated spatial Poisson example

We consider an outcome $y_{it}$ observed at 16 spatial locations and
192 time points. Two high-frequency covariates, $x_{1it}$ and $x_{2it}$,
are available for each location. There are 30 high-frequency observations
corresponding to each response time point.

The first covariate uses a hyperbolic lag-weight constraint with
$\gamma = 0.9$ and 29 lags. Its regression coefficient varies spatially
according to an intrinsic conditional autoregressive (iCAR) model.

The second covariate uses a Gaussian lag-weight constraint with
$\gamma_1 = 10$ and $\sqrt{\gamma_2} = 12$ and 45 lags. Its regression
coefficient is constant across locations.

The simulated data are included with the package and can be loaded using:

```{r load-data}
data("data_spatialpoisson_example")
```

The dataset is a list containing the response, the two high-frequency
covariates, and the spatial neighbourhood structure. For example:

```{r inspect-data}
names(data_spatialpoisson_example)
```

The response data contain the outcome, location index, and time index:

```{r response-data}
head(data_spatialpoisson_example[["data_y"]])
```

The high-frequency covariates contain the covariate values and their
corresponding location indices:

```{r covariate-data}
head(data_spatialpoisson_example$data_x1)
head(data_spatialpoisson_example$data_x2)
```

The neighbourhood structure used to generate the spatially varying
coefficient is stored as an inla.graph object:

```{r graph}
g <- INLA::inla.read.graph(
  filename = system.file("map.adj", package = "midasINLA")
)
```

The data were generated using a Poisson model of the form

$$
y_{it} \sim \operatorname{Poisson}(\mu_{it}),
$$

with

$$
\log(\mu_{it}) =
\beta_0 +
(\beta_1^* + b_i)
\sum_{k=0}^{29} w_{1k}x_{1i,s(t)-k}
+
\beta_2
\sum_{k=0}^{45} w_{2k}x_{2i,s(t)-k},
$$

where $\boldsymbol{b}$ follows an iCAR model. The true values of the
parameters are $\beta_0 = 1$, $\beta_1^* = 1.1$, and $\beta_2 = -2$.


## Preparing the MIDAS covariates

Before fitting the model, each high-frequency covariate is prepared using
`prepare_Minla_spatial`().

For the first covariate, we use the hyperbolic constraint and allow the
coefficient to vary spatially according to an iCAR model:


```{r prepare-x1}
Midas_x1 <- prepare_Minla_spatial(
  x = data_spatialpoisson_example$data_x1$x1,
  loc_x = data_spatialpoisson_example$data_x1$loc,
  constraint = "hyperbolic",
  K = 0:29,
  m = 30,
  svc = TRUE,
  svc_prior = "icar",
  g = g
)
```

```{r prepare-x2}
Midas_x2 <- prepare_Minla_spatial(
  x = data_spatialpoisson_example$data_x2$x2,
  loc_x = data_spatialpoisson_example$data_x2$loc,
  constraint = "gaussian",
  K = 0:45,
  m = 30,
  svc = FALSE
)
```

The resulting objects contain the information required by
`fit_Minla_spatial`() to construct the MIDAS components of the model.


## Fitting the model

The response data are stored in data_y. To illustrate prediction, we
reserve the final 10 response time points at each location as a test set.

```{r prepare-response}
response_data <- data_spatialpoisson_example[["data_y"]]

response_data$y_all <- response_data$y

response_data[which(response_data[["Time"]] %in% 183:192),"y"] <- NA
```

The model can then be fitted using `fit_Minla_spatial`():

```{r fit-model, eval = TRUE}
fit_res <- fit_Minla_spatial(
  formula = y ~ 1,
  data = response_data,
  loc_var = "loc",
  time_var = "Time",
  family = "poisson",
  hf_input = list(Midas_x1, Midas_x2),
  inla_options = list(verbose = FALSE)
)
```


A summary of the fitted model can be obtained using the standard
`summary`() method:

```{r model-summary}
summary(fit_res[["res"]])
```

The fitted model contains two MIDAS components, corresponding to the two
high-frequency covariates. The model also estimates the spatially varying
coefficient associated with the first covariate.


## Posterior summaries of regression coefficients

Posterior summaries of the MIDAS regression coefficients can be obtained
using `compute_beta_spatial`().

```{r beta-results}
beta_results <- compute_beta_spatial(
  model = fit_res,
  n_loc = 16
)
```

The returned object contains results for each high-frequency covariate.
For the first covariate, the posterior summaries of the spatially varying
component can be accessed using:

```{r beta-summary}
beta_results$hf_index_1$summary.icar.beta
```

The spatially varying coefficient at location $i$ is given by

$$
\beta_{1,i} = \beta_1^* + b_i.
$$

The posterior estimates can therefore be compared with the known
coefficients used to generate the simulated data.


## Estimating the MIDAS lag weights

The posterior distributions of the lag weights can be obtained using
`compute_weights`():

```{r weights}
res_weights <- compute_weights(fit_res)
```

The result is a list containing one data frame for each high-frequency
covariate. Each data frame contains the lag, posterior mean, and lower and
upper posterior quantiles.

For example, the estimated weights for the first covariate are:

```{r weights-x1}
head(res_weights$hf_1)
```

The weights for the second covariate can be inspected similarly:

```{r weights-x2}
head(res_weights$hf_2)
```

These posterior summaries can be used to visualise the estimated
lag-weight functions. For example:

```{r plot-weights, fig.width = 7, fig.height = 4.5}
plot(
  res_weights$hf_1$lag,
  res_weights$hf_1$mean,
  type = "l",
  xlab = "Lag",
  ylab = "Posterior mean weight",
  ylim = range(
    c(res_weights$hf_1$q2.5, res_weights$hf_1$q97.5)
  )
)

lines(
  res_weights$hf_1$lag,
  res_weights$hf_1$q2.5,
  lty = 2
)

lines(
  res_weights$hf_1$lag,
  res_weights$hf_1$q97.5,
  lty = 2
)
```

## Prediction for held-out observations

The final 10 time points were withheld from the model fit and can be used
to illustrate prediction.

Posterior predictions are generated using `predict_midas`():

```{r prediction}
pred_res <- predict_midas(
  model = fit_res,
  family = "poisson",
  Ntrials = NULL,
  nsamples = 1000
)
```

The returned object contains posterior summaries of the predicted outcome
as well as posterior samples of the latent predictor:

```{r prediction-structure}
str(pred_res, max.level = 2)
```


```{r prediction-summary}
head(pred_res$computed_y$mean)
head(pred_res$computed_y$q2.5)
head(pred_res$computed_y$q97.5)
```

The predictions can then be compared with the observations in the held-out
test set. For example, the following code displays the observed and
predicted values for the first location:

```{r plot-predictions, fig.width = 7, fig.height = 4.5}
loc1 <- response_data$loc == 1

time1 <- response_data$Time[loc1]
observed1 <- response_data$y_all[loc1]

pred_mean1 <- pred_res$computed_y$mean[loc1]
pred_lower1 <- pred_res$computed_y$q2.5[loc1]
pred_upper1 <- pred_res$computed_y$q97.5[loc1]

plot(
  time1,
  observed1,
  type = "p",
  xlab = "Time",
  ylab = "Outcome",
  pch = 16
)

lines(
  time1,
  pred_mean1,
  lty = 1
)

lines(
  time1,
  pred_lower1,
  lty = 2
)

lines(
  time1,
  pred_upper1,
  lty = 2
)

abline(
  v = 182.5,
  lty = 3
)
```

The vertical line separates the training period from the forecast test
period. The same approach can be used to obtain predictions for all
locations and to evaluate predictive performance using appropriate
pointwise or probabilistic scoring rules.


## Summary 

This vignette demonstrated the main workflow for fitting a spatial
SDL-MIDAS model using midasINLA.

The analysis consisted of the following steps:

1. Load the simulated spatial Poisson data.
2. Prepare each high-frequency covariate using `prepare_Minla_spatial`().
3. Fit the model using `fit_Minla_spatial`().
4. Obtain posterior summaries of the regression coefficients using
    `compute_beta_spatial`().
5. Obtain posterior summaries of the MIDAS lag weights using
    `compute_weights`().
6. Generate posterior predictions using `predict_midas`().

The simulated example is designed to illustrate the functionality of the
package. A real-data application demonstrating the use of midasINLA in a
substantive setting is presented in the accompanying R Journal article.
